---
title: "Final Project Exploration"
author: "Luis Calleja"
date: "December 16, 2018"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r, message = FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(stringr)
library(data.table)
library(ggplot2)
library(lubridate)
library(zoo)
library(forecast)
library(nlme)
library(skimr)
```

## MTA data
```{r}
subway <- read.csv('/home/lechuza/Documents/CUNY/data_621/finalProject/MTA_Performance_Agencies.csv', stringsAsFactors = FALSE)

str(subway)
```

Ride hailing data
```{r}
#For Hire Vehicle data
rh <- read.csv('/home/lechuza/Documents/CUNY/data_621/finalProject/rideHailing/FHV_Base_Aggregate_Report.csv', stringsAsFactors = FALSE)

str(rh)
```
Iden's code to join the target variables.
```{r}
rh %>%
  filter(grepl('uber|UBER|lyft|LYFT', rh$Base.Name)) -> ma

dt <- data.table(ma)

test <- dt[,sum(Total.Dispatched.Trips),by = c('Year','Month')]

names(test)[3] <- 'total.dispatched.trips'

#prep the MTA data:
subway <- subway[, c(4, 11, 12, 16)]

mdf <- subway %>%
  filter(INDICATOR_NAME == "Mean Distance Between Failures - Subways") %>%
  spread(INDICATOR_NAME, MONTHLY_ACTUAL)
tr <- subway %>%
  filter(INDICATOR_NAME == "Total Ridership - Subways") %>%
  spread(INDICATOR_NAME, MONTHLY_ACTUAL)
cir <- subway %>%
  filter(INDICATOR_NAME == "Customer Injury Rate - Subways") %>%
  spread(INDICATOR_NAME, MONTHLY_ACTUAL)
elev <- subway %>%
  filter(INDICATOR_NAME == "Elevator Availability - Subways") %>%
  spread(INDICATOR_NAME, MONTHLY_ACTUAL)
esc <- subway %>%
  filter(INDICATOR_NAME == "Escalator Availability - Subways") %>%
  spread(INDICATOR_NAME, MONTHLY_ACTUAL)
otp <- subway %>%
  filter(INDICATOR_NAME == "On-Time Performance (Terminal)") %>%
  spread(INDICATOR_NAME, MONTHLY_ACTUAL)
wait <- subway %>%
  filter(INDICATOR_NAME == "Subway Wait Assessment ") %>%
  spread(INDICATOR_NAME, MONTHLY_ACTUAL)

final.subway <- left_join(mdf, tr, by = c("PERIOD_YEAR", "PERIOD_MONTH"))
final.subway <- left_join(final.subway, cir, by = c("PERIOD_YEAR", "PERIOD_MONTH"))
final.subway <- left_join(final.subway, elev, by = c("PERIOD_YEAR", "PERIOD_MONTH"))
final.subway <- left_join(final.subway, esc, by = c("PERIOD_YEAR", "PERIOD_MONTH"))
final.subway <- left_join(final.subway, otp, by = c("PERIOD_YEAR", "PERIOD_MONTH"))
final.subway <- left_join(final.subway, wait, by = c("PERIOD_YEAR", "PERIOD_MONTH"))
```

Iden renames the dataframe and finalizes it.
```{r}
final.subway <- final.subway[-c(1:17),]
colnames(final.subway) <- c("YEAR", "MONTH", "FAILURE", "RIDERSHIP", "INJURY", "ELEV", "ESCA", "OTP","WAIT")

#merge the FHV to MTA
names(final.subway)
dt.fs <- data.table(final.subway)
```

Merge MTA dataset to the FHV data.
```{r}
tot <- merge(test,dt.fs, by.x = c('Year','Month'), by.y = c("YEAR","MONTH"), suffixes = c(".fhv",".mta"))
#36 observations... is that expected? Yes, the MTA data doesn't include 2018
tot.m <- merge(test,dt.fs, by.x = c('Year','Month'), by.y = c("YEAR","MONTH"), suffixes = c(".fhv",".mta"), all.x = TRUE)

names(tot.m)
z <- zoo::as.yearmon(paste(tot.m$Year, tot.m$Month, rep('01', length(tot.m$Year)),sep = '-'))

tot.m$month.date <- z
```

Overlay monthly public ridership, FHV ridership, mta on time performance
```{r}
par(mfrow = c(3,2))
plot.zoo(tot.m$month.date, tot.m$RIDERSHIP, type = 'l', main = "Ridership")
plot.zoo(tot.m$month.date, tot.m$FAILURE, type = 'l', main ="failures")
plot.zoo(tot.m$month.date, tot.m$OTP, type = 'l', main = "On time performance")
plot.zoo(tot.m$month.date, tot.m$total.dispatched.trips, type = 'l', main = "FHV Dispatched")
plot.zoo(tot.m$month.date, tot.m$WAIT, type = 'l', main = "Subway Wait Assessment")
```

Analyze seasonality and trend for the response and predictor variables as they are all temporal.
```{r}
trend_ridership <- ma(tot.m$RIDERSHIP, order = 12, centre = TRUE)
par(mfrow = c(2,1))
plot(as.ts(tot.m$RIDERSHIP))
lines(trend_ridership)
plot(as.ts(trend_ridership))

ts_ride <- ts(tot.m$RIDERSHIP, frequency = 12)
ts.fhv <- ts(tot.m$total.dispatched.trips, frequency = 12)
ts.fail <- ts(tot.m$FAILURE, frequency = 12)
ts.otp <- ts(tot.m$OTP, frequency = 12)
ts.wait <- ts(tot.m$WAIT, frequency = 12)

ts.ride.de <- decompose(ts_ride, "multiplicative")
ts.fhv.de <- decompose(ts.fhv, "multiplicative")
ts.fail.de <- decompose(ts.fail, "multiplicative")
ts.otp.de <- decompose(ts.otp, "multiplicative")

#plot ridership deconstructed time series
plot(decompose(ts_ride, "multiplicative"))
#plot ride hailing dispatched trips deconstructed time series
plot(decompose(ts.fhv, "multiplicative"))
#plot failure MTA deconstructed time series
plot(decompose(ts.fail, "multiplicative"))
#plot on time performance deconstructed time series
plot(decompose(ts.otp, "multiplicative"))
```

Look at the autocorrelation tendencies from each variable.
```{r}
head(ts.ride.de$random)
#acf(ts.ride.de$random)
par(mfrow=c(1,1))
acf(tot.m$RIDERSHIP[!is.na(tot.m$RIDERSHIP)])
```
This confirms the seasonality inherent in the ridership time series.

Plot the autocorrelation measure among the predictors.
```{r}
acf(tot.m$total.dispatched.trips[!is.na(tot.m$total.dispatched.trips)])
acf(tot.m$OTP[!is.na(tot.m$OTP)])
acf(tot.m$WAIT[!is.na(tot.m$WAIT)])
```

Initially... my desire was to build two models:
1) The first model is built using the entirely decomposed time series with trend and seasonality removed.
2) Second model is built using the trend datasets (with seasonality removed) from each variable.

This is not very feasible as the de-trending and seasonality removal would leave us too few observations (<30). Instead, we'll fit a more naive lm and account for any serial correlation via the GLS approach.

Plot a time series model on lags of first and second order of each of the predictors.
```{r}
tot.m %>%
  dplyr::select(-INJURY) -> temp

temp[complete.cases(temp)] -> tot.m.cc

lagged.set = data.frame(
  ridership = lag(tot.m.cc$RIDERSHIP,1),
fhv = tot.m.cc$total.dispatched.trips,
failure = tot.m.cc$FAILURE,
otp = tot.m.cc$OTP,
wait = tot.m.cc$WAIT,
month = tot.m.cc$Month)

lagged.set <- lagged.set[complete.cases(lagged.set),]
```

Fit a model on the lagged predictors
```{r}
model.ts <- lm(data = lagged.set, ridership ~ .)

summary(model.ts)
```

Wow, the model is hot garbage!

Model diagnostics
```{r}
par(mfrow=c(1,1))
plot(model.ts$fitted, rstandard(model.ts))

qqnorm(residuals(model.ts))
qqline(residuals(model.ts))
```

The qqplot demonstrates a pattern among the residuals, suggesting the model is not a good fit.

Investigate whether there is any autocorrelation among the error terms.
```{r}
acf(model.ts$residuals)
```
There is an indication of lag 12 serial correlation of the errors.

Treat the response variable, then re-fit.
```{r}
model.ts.log <- lm(data = lagged.set, I(log(ridership) ~ .))

summary(model.ts.log)
```

Fit a generalized least squares model
```{r}
gls.model <-gls(data = lagged.set, ridership ~ .)

residuals(gls.model)
```

Treat the response and all predictor variables, then re-fit.

We'll fit the model on the month-over-month growth of each of the variables.
```{r}
dt.l.s <- data.table(lagged.set)

testy <- dt.l.s[,lapply(.SD,function(x) {diff(log(x),1)})]

testy[,month:= lagged.set$month[1:34]]
```

```{r}
model.ts.log <- lm(data = testy, ridership ~ .)

summary(model.ts.log)
```

Iden's model 
```{r}
fs.cc <- final.subway[complete.cases(final.subway),]
model <- lm(OTP ~ FAILURE + RIDERSHIP + INJURY + ELEV + ESCA, data = fs.cc)

summary(model)
```
Plot the residuals... looks like there is a pattern of the residuals... almost a negative quadratic
```{r}
par(mfrow= c(1,1))
plot(model$fitted, rstandard(model))
```

Plot each variable against the residuals
```{r}
fs.cc %>%
  dplyr::select(FAILURE, RIDERSHIP, ELEV, ESCA, INJURY) %>%
  gather() -> long.form

long.form$errors <- rep(rstandard(model),5)

ggplot(long.form,aes(x = value, y = errors)) +geom_point() + facet_wrap(~key, scales = "free")
```